Verification of concurrent design patterns with data

### 2019-02-15 2018-Arch-pNets/arch-pNets.tex
3 sec in 9 pages

Wwe propose   to encode BIP architectures
into open pNet and to verify, using SMT techniques,
the coordination capabilities of the architecture, and 
its properties. The encoding will  give us feedback on the
comparison between the BIP architectures and the pNet formalism.  
Overall, the encoding allows us to precisely compare the coordination
mechanisms of the two models but also the way component composition is
expressed in the two formalisms. This  also gives the opportunity
to
% investigate further on  compositional verification by
relate two  different
visions on compositional verification. 
%\noteLH{pas tres heureux la fin:
%  pas assez detaille ou trop complique -> abstraire ou detailller a
%  mon avis (Ludo). J'ai prefere abstraire/raccourcir... (Eric)}

%\paragraph{Contributions}
The main contributions of this paper are: 1) The definition of a
theory of BIP architectures with data, including a theorem about
preservation of (data dependent) properties by architecture
compositions. 2) An encoding of architectures with data into the open pNet
model, allowing for analysis of their temporal properties using pNet's
software tools.
The concepts and results of the paper are illustrated by a running
example extending the Failure Monitor architecture from the CubETH
nanosatellite on-board software case study~\cite{CubETH-case-study}.

@@

\subsection{Notations}
\label{secn:notations}

@@

For a set of variables $V \subseteq \variables$, we denote
$\valuations{V} \bydef{=} \{\val{}{}: V \rightarrow \data\}$
the set of \emph{valuations} of the variables in $V$.
Valuations extend canonically to expressions and denote $\val{}{}(e)$ the value of $e$ under the valuation $\val{}{}$.
%\noteLH{its the following really necessary}
%\addSB{Since we consider variables to be sorted, valuations
%  applied to expressions are interpreted as partial functions
%  $\exprs{V} \rightharpoonup \data$.}
For a valuation $\val{}{}$ and an assignment $(x_i := e_i)^{i \in I}$, we define:
\vspace{-1.2ex}
\[
\val{}{}\bigl[(x_i := e_i)^{i \in I}\bigr](x) \bydef{=}
\begin{cases}
  \val{}{}(x), & \text{if } x \not\in x_i^{i \in I}\,,\\
  \val{}{}(e_i), & \text{if } x = x_i, \text{for some } i \in I\,.
\end{cases}
\]
For two valuations $\val{1}{}, \val{2}{} : V \rightarrow
\data$, we denote $\valdiff{\val{1}{}}{\val{2}{}} \bydef{=}
\bsetdef{x \in V}{\val{1}{}(x) \neq \val{2}{}(x)}$ the set of
variables that are assigned different values by the two
valuations.  As usual, we write $\val{1}{} \order \val{2}{}$
iff $\val{1}{}(x) \order \val{2}{}(x)$, for all $x \in V$.  An
expression $e$ is  \emph{monotonic} if, for
any two valuations $\val{1}{}, \val{2}{}$,
$\val{1}{} \order \val{2}{}$ implies $\val{1}{}(e) \order
\val{2}{}(e)$.
%
Similarly, an assignment $(x_i := e_i)^{i \in I} $ is
 monotonic if all expressions $e_i^{i \in I}$ are monotonic.
We  denote $\guards{V} \subset \boolexprs{V}$, $\monexprs{V}
\subset \exprs{V}$ and $\updates{V} \subset \assigns{V}$ the sets of
monotonic Boolean and generic expressions and assignments,
respectively.
%
%****************************************************************

\subsection{Open pNets}
\label{secn:pNets}
This section briefly describes pNets, see~\cite{HMZ-FORTE2016} for more complete   description.
pNets are tree-like structures, where the leaves are either
\emph{parameterised labelled transition systems (pLTSs)}, expressing the
behaviour of basic processes, or \emph{holes}, used as placeholders
for unknown processes.
Nodes of the tree are synchronising artefacts using a
set of \emph{synchronisation vectors} that express the possible
synchronisation between the parameterised actions of a subset of the
sub-trees.

%\paragraph{Definitions.}
%\label{section:pNets}
A pLTS is a labelled transition system with variables that can be
used to characterise a state, as part of actions, guards, and
assignments. 
%Without loss of generality and to simplify the formalisation, we suppose 
%here that 
%variables are local to each 
%state: each state has its set of variables disjoint from the others. Transmitting 
%variable values from one state to the other can be done by explicit assignment. 
%Similarly, to simplify the management of variables and without loss of expressivity, we 
%suppose that transitions looping to the same state does not do assignments.
%Note that we make no assumption on the finiteness of the set of states, nor
%on the finite branching of the transition relation.
We need to introduce input variables to reason about the data flow between actions.
We  define the set of actions a pLTS can use ($a$
ranges over action labels):\\
% Action terms are  of the form 
$\alpha=a(?x_i^{i \in I}, 
\Expr_j^{j \in J})$, where $?x_i^{i \in I}$ are input variables, $\Expr_j^{j \in J}$ 
are expressions.

@@

\begin{definition}[pNets]\label{defn:pnets}
A pNet is a hierarchical structure where leaves are pLTSs and holes: %\footnote{Typing can 
%be added to pNets by defining and checking their interfaces, called 
%sort~\cite{HMZ-FORTE2016}}:
$\pNet\triangleq \pLTS~|~\mylangle \pNet_i^{i\in I}, J, \symb{SV}_k^{k\in 
K}\myrangle$
where\\[-3.5ex]
\begin{itemize}
\item $\pNet_i^{i\in I}$ is the family of sub-pNets;
%  $\pNet_i^{i\in I}$ is a family of sub-pNets where $I\in\I_\P$ is the set over which 
%sub-pNets are indexed.

@@

\end{itemize}
%We define the set of holes and leaves of a pNet as follows:
%  \begin{itemize}
%\item
The set of \emph{holes} $\Holes(\pNet)$ of a pNet is the indexes of the holes of the pNet 
itself plus the indexes of all the holes of its subnets (we suppose those indexes 
disjoints). A pNet $Q$ is \emph{closed} if it has no hole: $\Holes(Q)=\emptyset$; else it
is said to be \emph{open}.
%
%defined inductively; the sets of holes
%in a pNet node and its subnets are all disjoint:
%  \[\begin{array}{l}
%\Holes(\mylangle S,s_0, \to\myrangle) \!=\! \emptyset \\
%\Holes(\mylangle \pNet_i^{i\in I}\!,\Sort_j^{j\in J}\!, \overline{\symb{SV}}\myrangle) 
%=J\uplus{\displaystyle \bigcup_{i\in 
%I}\Holes(\pNet_i)}\\
%\forall i\in I.\, \Holes(\pNet_i)\cap J=\emptyset\\
%\forall i_1,i_2\in I.\,i_1\neq i_2\Rightarrow  
%\Holes(\pNet_{i_1})\cap\Holes(\pNet_{i_2})=\emptyset
%\end{array}\]
%\item
The set of \emph{leaves} of a pNet is the set of all pLTSs occurring in the structure, as an 
indexed family of the form $\Leaves(\pNet)= \mylangle \pNet_i \myrangle^{i \in L}$.
%\end{itemize}

@@

The semantics of an open pNets is an open automaton where  the states are
tuples of states of the pLTSs at the leaves, denoted $\triangleleft\ldots\triangleright$.  Each open transition between two states is defined as a composition of transition from some pLTSs and actions of some holes as follows: 1) the actions of the holes involved in the transition, 2) a guard  built from the synchronisation vectors coordinating the holes and the transitions involved; 3) assignments and global state change are defined by the pLTSs transitions involved. The synchronisation vectors also define a global action.

@@

With pNets, we can define a behavioural semantics for distributed components~\cite{AmeurBoulifa2017} where correctness properties are verified by model-checking. 
More recently, a bisimulation theory has been formalised for open pNets~\cite{HMZ-FORTE2016}.

@@

>>>>

We propose   to encode BIP architectures
into open pNet and to verify, using SMT techniques,
the properties of the architectures. The encoding also allows us to
compare BIP architectures and  pNet formalism.  
Overall, the encoding allows us to precisely compare the coordination
mechanisms of the two models but also the way component composition is
expressed in the two formalisms. This  also gives the opportunity
to
% investigate further on  compositional verification by
relate two  different
visions on compositional verification. 
%\noteLH{pas tres heureux la fin:
%  pas assez detaille ou trop complique -> abstraire ou detailller a
%  mon avis (Ludo). J'ai prefere abstraire/raccourcir... (Eric)}

%\paragraph{Contributions}
The main contributions of this paper are: 1) The definition of a
theory of BIP architectures with data, including a theorem about
preservation of (data dependent) properties by 
compositions. 2) An encoding of architectures with data into  open pNets, allowing for analysis of their temporal properties using pNet's
software tools.
The  paper is illustrated by a running
example based on the failure monitor architecture from the CubETH
nanosatellite on-board software~\cite{CubETH-case-study}.

@@

\subsubsection{Notations.}
\label{secn:notations}

@@

For a set of variables $V \subseteq \variables$, we denote
$\valuations{V} \bydef{=} \{\val{}{}: V \rightarrow \data\}$
the set of \emph{valuations} of the variables in $V$ and let $\sigma$ range over valuations.
Valuations extend canonically to expressions, denoted $\val{}{}(e)$. We define:
%\noteLH{its the following really necessary}
%\addSB{Since we consider variables to be sorted, valuations
%  applied to expressions are interpreted as partial functions
%  $\exprs{V} \rightharpoonup \data$.}
%For a valuation $\val{}{}$ and an assignment $(x_i := e_i)^{i \in I}$, we define:
\vspace{-1.5ex}
\[
\val{}{}\bigl[(x_i := e_i)^{i \in I}\bigr](x) \bydef{=}
\begin{cases}
  \val{}{}(x), & \text{if } x \not\in x_i^{i \in I}\,,\\
  \val{}{}(e_i), & \text{if } x = x_i, \text{for some } i \in I\,.
\end{cases}\vspace{-1ex}
\]
For two valuations $\val{1}{}, \val{2}{} : V \rightarrow
\data$, we denote $\valdiff{\val{1}{}}{\val{2}{}} \bydef{=}
\bsetdef{x \in V}{\val{1}{}(x) \neq \val{2}{}(x)}$ the set of
variables that are assigned different values by the two
valuations.  As usual, we write $\val{1}{} \order \val{2}{}$
iff $\val{1}{}(x) \order \val{2}{}(x)$, for all $x \in V$.  An
expression $e$ is  \emph{monotonic} if, for
any two valuations $\val{1}{}, \val{2}{}$,
$\val{1}{} \order \val{2}{}$ implies $\val{1}{}(e) \order
\val{2}{}(e)$.
%
Similarly, an assignment $(x_i := e_i)^{i \in I} $ is
 monotonic if all expressions $e_i^{i \in I}$ are monotonic.
We  denote $\guards{V} \subset \boolexprs{V}$, $\monexprs{V}
\subset \exprs{V}$ and $\updates{V} \subset \assigns{V}$ the sets of
monotonic Boolean and generic expressions and assignments,
respectively.
%
%****************************************************************

\subsubsection{Open pNets.}
\label{secn:pNets}
This section briefly describes pNets, see~\cite{HMZ-FORTE2016} for more complete   description.
pNets are tree-like structures, where the leaves are either
\emph{parameterised labelled transition systems (pLTSs)}, expressing the
behaviour of basic processes, or \emph{holes}, used as placeholders
for unknown processes.
Nodes of the tree are synchronising artefacts using a
set of \emph{synchronisation vectors} that express the possible
synchronisation between  parameterised actions of some components.

%\paragraph{Definitions.}
%\label{section:pNets}
A pLTS is a labelled transition system with variables that can be
used to characterise a state, as part of actions, guards, and
assignments. 
%Without loss of generality and to simplify the formalisation, we suppose 
%here that 
%variables are local to each 
%state: each state has its set of variables disjoint from the others. Transmitting 
%variable values from one state to the other can be done by explicit assignment. 
%Similarly, to simplify the management of variables and without loss of expressivity, we 
%suppose that transitions looping to the same state does not do assignments.
%Note that we make no assumption on the finiteness of the set of states, nor
%on the finite branching of the transition relation.
We need to introduce input variables to reason about the data flow between actions.
We  define the set of actions a pLTS can use ($a$
ranges over action labels):\\
% Action terms are  of the form 
$\alpha=a(?x_i^{i \in I}, 
e_j^{j \in J})$, where $?x_i^{i \in I}$ are input variables, $e_j^{j \in J}$ 
are expressions.

@@

\begin{definition}[pNets]\label{defn:pnets}
A pNet is a hierarchical structure where leaves are pLTSs and holes: %\footnote{Typing can 
%be added to pNets by defining and checking their interfaces, called 
%sort~\cite{HMZ-FORTE2016}}:
$Q\triangleq \pLTS~|~\mylangle Q_i^{i\in I}, J, \symb{SV}_k^{k\in 
K}\myrangle$
where\\[-3.5ex]
\begin{itemize}
\item $Q_i^{i\in I}$ is the family of sub-pNets;
%  $Q_i^{i\in I}$ is a family of sub-pNets where $I\in\I_\P$ is the set over which 
%sub-pNets are indexed.

@@

\end{itemize}
%We define the set of holes and leaves of a pNet as follows:
%  \begin{itemize}
%\item
The set of \emph{holes} $\Holes(Q)$ of a pNet is the indexes of the holes of the pNet 
itself plus the indexes of all the holes of its subnets (we suppose those indexes 
disjoints). A pNet $Q$ is \emph{closed} if it has no hole: $\Holes(Q)=\emptyset$; else it
is said to be \emph{open}.
%
%defined inductively; the sets of holes
%in a pNet node and its subnets are all disjoint:
%  \[\begin{array}{l}
%\Holes(\mylangle S,s_0, \to\myrangle) \!=\! \emptyset \\
%\Holes(\mylangle \pNet_i^{i\in I}\!,\Sort_j^{j\in J}\!, \overline{\symb{SV}}\myrangle) 
%=J\uplus{\displaystyle \bigcup_{i\in 
%I}\Holes(\pNet_i)}\\
%\forall i\in I.\, \Holes(\pNet_i)\cap J=\emptyset\\
%\forall i_1,i_2\in I.\,i_1\neq i_2\Rightarrow  
%\Holes(\pNet_{i_1})\cap\Holes(\pNet_{i_2})=\emptyset
%\end{array}\]
%\item
The set of \emph{leaves} of a pNet is the set of all pLTSs occurring in the structure, as an 
indexed family of the form $\Leaves(Q)= \mylangle \pLTS_i \myrangle^{i \in L}$.
%\end{itemize}

@@

The semantics of an open pNets is an open automaton where  the states are
tuples of states of the pLTSs at the leaves, denoted $\triangleleft\ldots\triangleright$.  Each open transition between two states contains 1) the actions of the holes involved in the transition, 2) a guard  built from the synchronisation vectors coordinating the holes and the transitions involved; 3) assignments and global state change  defined by the pLTSs transitions involved; 4) a global action  defined by the synchronisation vector.

@@

@@

We  used pNets to define a behavioural semantics for distributed components~\cite{AmeurBoulifa2017} that allows the verification of correctness properties  by model-checking. 
More recently, a bisimulation theory has been formalised for open pNets~\cite{HMZ-FORTE2016}.
